### 2018-02-19 submission/manuscript.Rmd
update abstract with edits

**Background.** An increasing body of literature suggests that there is a crucial role for the microbiota in colorectal cancer (CRC) pathogenesis. Important drivers within this context have ranged from individual microbes to the whole community. Our study expands on a recent meta-analysis investigating microbial biomarkers for tumors by testing the hypothesis that the bacterial community has important associations to both adenoma and carcinoma tumors. To test this hypothesis we examined both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) across 14 previously published 16S rRNA gene sequencing studies on colorectal tumors and the microbiota.

**Results.** Fecal samples had a significant decrease for both Shannon diversity and evenness as tumor severity increased, after correcting for study effect and variable region sequenced (P-value `r ifelse(min(filter(stool_lme, !(alpha_metric %in% "sobs"))[, "BH"]) < 0.05, invisible("<"), invisible("check tables"))` 0.05). This reduction in evenness translated into small increases in the odds ratio for individuals to have both adenoma (P-value = `r format(min(filter(alpha_RR_adn_stool, !(measure %in% "sobs"))[, "pvalue"]), digits = 2)`) and carcinoma tumors (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2)`) while the reduction in Shannon diversity only translated into an increased odds ratio for individuals to have carcinomas (P-value = `r format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2)`). Increases in mouth-associated microbes were commonly in the top 5 most increased odds ratios for individuals to have either adenoma or carcinoma tumors, regardless of sample type. Prediction models built to classify either individuals with adenoma or carcinoma were trained on the whole community or selected genera (top 5 highest and lowest odds ratios) from either fecal or tissue samples. Both the full and select models for either adenoma or carcinoma resulted in similar classification success according to Area Under the Curve (AUC). The most important groups within the full community models consistently belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* across studies. Although a number of associations between the microbiota and tumor were identified, the majority of studies that we used in this meta-analysis were only individually adequately powered for large effect sizes.

**Conclusions.** These data provide support for the importance of the bacterial community to both adenoma and carcinoma tumorgenesis. The evidence collected within this study on the role of the microbiota in those with tumors identifies a number of correlations that may not have been detected because of the low power associated with the majority of studies that have been performed to date.

>>>>

**Background.** An increasing body of literature suggests that both individual and collections of bacteria are associated with the progression of colorectal cancer. As the number of studies investigating these associations increases and the number of subjects in each study increases, a meta-analysis to identify the associations that are the most predictive of disease progression is warranted. For our meta-analysis, we analyzed previously published 16S rRNA gene sequencing data collected from feces (`r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)` individuals from `r length(stool_n$study)` studies) and colon tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals from `r length(tissue_n$study)` studies).

**Results.** We quantified the odds ratios for individual bacterial genera that were associated with an individual having tumors relative to a normal colon. Among the stool samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_OR_stool$measure)` genera with significant odds ratios associated with carcinoma. Similarly, among the tissue samples, there were no genera that had a significant odds ratio associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` genera with significant odds ratios associated with carcinoma. Among the significant odds ratios, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`. Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using the genera with the five highest and lowest odds ratios, using the entire collection of genera found in each study, and using operational taxonomic units defined based on a 97% similarity threshold. All training approaches yielded similar classification success as measured using the  Area Under the Curve. The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from stool or tissue.

**Conclusions.** This meta-analysis confirms previous results indicating that individuals with adenomas cannot be readily classified based on their bacterial community, but that those with carcinomas can. Regardless of the dataset, we found a subset of the fecal community that was associated with carcinomas was as predictive as the full community.

### 2018-02-19 submission/manuscript.tex
update abstract with edits

\textbf{Background.} An increasing body of literature suggests that
there is a crucial role for the microbiota in colorectal cancer (CRC)
pathogenesis. Important drivers within this context have ranged from
individual microbes to the whole community. Our study expands on a
recent meta-analysis investigating microbial biomarkers for tumors by
testing the hypothesis that the bacterial community has important
associations to both adenoma and carcinoma tumors. To test this
hypothesis we examined both feces (n = 1737) and colon tissue (492 total
samples from 350 individuals) across 14 previously published 16S rRNA
gene sequencing studies on colorectal tumors and the microbiota.

\textbf{Results.} Fecal samples had a significant decrease for both
Shannon diversity and evenness as tumor severity increased, after
correcting for study effect and variable region sequenced (P-value
\textless{} 0.05). This reduction in evenness translated into small
increases in the odds ratio for individuals to have both adenoma
(P-value = 0.035) and carcinoma tumors (P-value = 0.0021) while the
reduction in Shannon diversity only translated into an increased odds
ratio for individuals to have carcinomas (P-value = 0.0069). Increases
in mouth-associated microbes were commonly in the top 5 most increased
odds ratios for individuals to have either adenoma or carcinoma tumors,
regardless of sample type. Prediction models built to classify either
individuals with adenoma or carcinoma were trained on the whole
community or selected genera (top 5 highest and lowest odds ratios) from
either fecal or tissue samples. Both the full and select models for
either adenoma or carcinoma resulted in similar classification success
according to Area Under the Curve (AUC). The most important groups
within the full community models consistently belonged to genera such as
\emph{Ruminococcus}, \emph{Bacteroides}, and \emph{Roseburia} across
studies. Although a number of associations between the microbiota and
tumor were identified, the majority of studies that we used in this
meta-analysis were only individually adequately powered for large effect
sizes.

\textbf{Conclusions.} These data provide support for the importance of
the bacterial community to both adenoma and carcinoma tumorgenesis. The
evidence collected within this study on the role of the microbiota in
those with tumors identifies a number of correlations that may not have
been detected because of the low power associated with the majority of
studies that have been performed to date.

>>>>

\textbf{Background.} An increasing body of literature suggests that both
individual and collections of bacteria are associated with the
progression of colorectal cancer. As the number of studies investigating
these associations increases and the number of subjects in each study
increases, a meta-analysis to identify the associations that are the
most predictive of disease progression is warranted. For our
meta-analysis, we analyzed previously published 16S rRNA gene sequencing
data collected from feces (1737 individuals from 8 studies) and colon
tissue (492 total samples from 350 individuals from 7 studies).

\textbf{Results.} We quantified the odds ratios for individual bacterial
genera that were associated with an individual having tumors relative to
a normal colon. Among the stool samples, there were no genera that had a
significant odds ratio associated with adenoma and there were 8 genera
with significant odds ratios associated with carcinoma. Similarly, among
the tissue samples, there were no genera that had a significant odds
ratio associated with adenoma and there were 3 genera with significant
odds ratios associated with carcinoma. Among the significant odds
ratios, the association between individual taxa and tumor diagnosis was
equal or below 7.11. Because individual taxa had limited association
with tumor diagnosis, we trained Random Forest classification models
using the genera with the five highest and lowest odds ratios, using the
entire collection of genera found in each study, and using operational
taxonomic units defined based on a 97\% similarity threshold. All
training approaches yielded similar classification success as measured
using the Area Under the Curve. The ability to correctly classify
individuals with adenomas was poor and the ability to classify
individuals with carcinomas was considerably better using sequences from
stool or tissue.

\textbf{Conclusions.} This meta-analysis confirms previous results
indicating that individuals with adenomas cannot be readily classified
based on their bacterial community, but that those with carcinomas can.
Regardless of the dataset, we found a subset of the fecal community that
was associated with carcinomas was as predictive as the full community.
